---
title: "Spotify analysis: Track duration of german rap music"
subtitle: "Part 1: Preprocessing"
author: "Luca Hüsler, Fabian Trottmann"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: pygments
    toc: true
    toc_float: true
    df_print: paged
    number_sections: true
# knit documents to output folder
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, cache = TRUE)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
```


# Load data

We first load the data that we extracted from the Spotify API via the *loadData.R* script.

```{r}
df.german_rap_features <- read_rds("../data/audio_features_by_german_hiphop_2020-02-21.rds")
df.german_rap_artists <- read_rds("../data/german_rap_artists_2020-02-21.rds")
```

# Inpsect the data
We do some basic checks to see if the import went well.
```{r}
dim(df.german_rap_features)
dim(df.german_rap_artists)
```

We obtain 7230 rows (tracks) with 40 variables. The artist data contains 50 rows with 12 variables.

Let's have a look at the first rows:
```{r}
head(df.german_rap_features)
```
Looks good. 

# Data cleaning
## Filterting & Cleaning
To clean the data, we rely upon **[tidyverse](https://www.tidyverse.org/)**, a collection of R packages designed for data science. We use piping to clean the data and transform it to the desired shape for the analysis. The pipe consists of the following steps:

- join potentially interesting variables of the artist data (popularity & number of followers) 
- select columns of interest for the analysis
- select distinct `track_name`'s (as features appear multiple times, for each artist)
- create a new column: `duration_min`
- filter out `track_name`'s that contain "intro", "outro", "skit" or "instrumental", as those records should not be considered for the analysis
- filter on `duration_min`: only tracks with a duration between 1 and 10 minutes will be considered (decision made by the authors to remove outliers)
- filter on `album_release_year`: only tracks released between 2015 and 2019 will be considered


```{r}
df.german_rap_features_clean <- df.german_rap_features %>%
  left_join(df.german_rap_artists, by = c("artist_id" = "id")) %>%
  select(track_id, track_name, track_popularity, artist_id, artist_name, duration_ms, album_release_date,
         danceability, energy, loudness, key, mode, speechiness, acousticness, liveness, instrumentalness, liveness,
         tempo, valence, explicit, key_name, mode_name, key_mode,
         album_release_year, album_type, album_name, key_mode, key_name, followers.total) %>%
  dplyr::distinct(track_id, .keep_all = TRUE) %>%
  dplyr::distinct(track_name, artist_name, .keep_all = TRUE) %>%
  filter(!str_detect(track_name, 
                     "Intro|Outro|Commentary|Skit|Instrumental")) %>%
  dplyr::mutate(duration_min = duration_ms/60000) %>%
  filter(duration_min > 1.00, duration_min < 10.00) %>%
  filter(album_release_year > 2014, album_release_year < 2020)

```


Let's check the data after preprocessing and cleaning.
```{r}
dim(df.german_rap_features_clean)
```

## Type conversions
As a next step, we check for the data types of the remaining variables and see if we have to do some type conversions for the analysis.
```{r}
str(df.german_rap_features_clean)
```

We should do some type conversions to make the variables suitable for analysis:

- change strings to factors for `key_name`, `mode_name`, `key_mode` and `album_type`
- change string to date from `album_release_year` (in case we consider that in the model)

```{r}
df.german_rap_features_clean <- df.german_rap_features_clean %>%
  mutate(album_release_date = lubridate::as_date(album_release_date)) %>%
  mutate_at(vars(key_name, mode_name, key_mode, album_type), factor)
```

Let's check if that worked out:
```{r}
df.german_rap_features_clean %>%
  select(album_release_date, key_name, mode_name, key_mode, album_type) %>%
  str()
```

Ok, data types look good now.

## Missing values
Finally, we should also check for missing values before entering the analysis. Let's get an overview by calculating the mean of NA's for each column:
```{r}
df.german_rap_features_clean %>%
  is.na() %>%
  colMeans() 
```
That looks perfect, no missing values. Ready for analysis.

# Output 
We can now write the preprocessed and cleaned dataframe to an .rds file which will be used for the analysis subsequently.

```{r}
write_rds(df.german_rap_features_clean, paste0("../data/german_rap_features_clean_", Sys.Date(), ".rds"))
```




